Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-27 RevisedSubmission/hines_hetland_2017_gji.tex
did more

We next establish our prior model for transient displacements.
Specifically, we discuss our choice for the covariance functions
$X(\pos,\pos')$ and $T(t,t')$. For $X$, we use the squared exponential
(SE) covariance function,
\begin{equation}\label{eq:SE}
X(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2 \ell^2}\right).
\end{equation}
The SE covariance function is commonly used in kriging
\citep[e.g,][]{Cressie1992} and Gaussian process regression
\citep[e.g.,][]{Rasmussen2006}. In terms of geodetic applications,
\citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE
accurately describes the covariance of secular GNSS derived velocities
in Japan. The SE is a positive definite covariance function for any
number of spatial dimensions. A Gaussian process with an SE covariance
function is isotropic and has realizations that are infinitely
differentiable.

We consider three potential models for the temporal covariance of $u$.
First, we consider the one-dimensional SE covariance function,
\begin{equation}\label{eq:TimeSE}
T(t,t') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right).
\end{equation}
Note that $T$ includes the hyperparameter $\phi$, which serves to
scale the covariance function $C_u$. Second, we consider integrated
Brownian motion (IBM). IBM has zero mean and its covariance function
can be found by integrating the covariance function for Brownian
motion as
\begin{align}\label{eq:IBM}
T(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(s,s') \,ds'\,ds \\
        &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used in the context of Kalman filtering as a
non-parametric model for the time dependence of geophysical signals
\citep[e.g.,][]{Segall1997, McGuire2003, Ohtani2010, Hines2016a}. It
should be emphasized $t=0$ is a reference time at which the Gaussian
process is exactly zero. For some geophysical signals, it is
appropriate to have this reference time. For example, if we are trying
to identify postseismic deformation then $t$ should be zero at the
time of the earthquake.  However, if we are interesting in detecting
transient events, where there is no known start time, then IBM may not
be an appropriate prior, and an isotropic Gaussian process should be
preferred. In the following analysis, we make the quite arbitrary
choice that $t$ is zero on the first epoch of $\data_*$. Using an
earlier reference time does not change the results discussed in this
section. Our third option for $T$ is the Wendland class of covariance
functions \citep{Wendland2005}. Wendland covariance functions have
compact support and hence their corresponding covariance matrices are
sparse. In our analysis, we exploit this sparsity with the CHOLMOD
software package \citep{Chen2008}. Wendland functions are positive
definite in $\mathbb{R}^d$, and they describes an isotropic Gaussian
process with realizations that can be differentiated $k$ times. The
form of the covariance function depends on the choice of $d$ and $k$.
We use $d=1$ since we are describing the temporal covariance of $u$.
We use $k=2$, giving samples of $u$ continuous velocities and
accelerations. The corresponding Wendland covariance function is
\begin{equation}\label{eq:Wendland}
T(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\tau}\right)^5_+ \left(\frac{8|t - t'|^2}{\tau^2} + \frac{5|t - t'|}{\tau} + 1\right), 
\end{equation}
where
\begin{equation}
(t)_+ = 
\begin{cases}
t, \ \ \ t > 0 \\
0, \ \ \ \mathrm{otherwise}.
\end{cases}
\end{equation}

We next determine appropriate hyperparameters for $X$ and each of the
three candidate covariance functions for $T$. we divide the dataset
into seven subsets which are four months long and each centered on the
time of a SSE. The times of the seven SSEs are determined with tremor
records from \cite{Wech2010}. We use the REML method to find the
optimal hyperparameters for $T$ and $X$ for each subset of data. We
choose to make each data subsets four months long because it is long
enough to encompass a SSE in the Pacific Northwest, while it is short
enough to still be computationally tractable. However, four months is
too short to resolve the sinusoids in $\data$, and they are omitted
from $\data$ in this REML analysis for Pacific Northwest SSEs. The
estimated hyperparameters for $u$ are summarized in Table 1. Based on
the interquartile ranges, the estimated hyperparameters for the SE and
Wendland covariance functions do not vary significantly between SSEs.
This suggests that the median of estimated hyperparameters should be
an appropriate prior model for all Pacific Northwest SSEs. For the IBM
model, there are several anomalously large values for $\ell$ and
$\phi$, which results in large interquartile ranges.

>>>>

% State our choice for X_i

We next establish our prior model for transient displacements.
Specifically, we discuss our choice for the functions $X_i$ and $T_i$.
For $X_i$, we use the squared exponential (SE) function,
\begin{equation}\label{eq:SE}
X_i(\pos,\pos') = \exp\left(\frac{-||\pos - \pos'||_2^2}{2 \ell^2}\right).
\end{equation}
As we have mentioned, the SE function is commonly used for kriging
\citep[e.g,][]{Cressie1992} and Gaussian process regression
\citep[e.g.,][]{Rasmussen2006}. In terms of geodetic applications,
\citet{Kato1998} and \cite{El-Fiky1999} demonstrated that the SE
function accurately describes the spatial covariance of secular GNSS
derived velocities in Japan.

% List our choices for T_i

We consider three potential choices for $T_i$. The first option is the
one-dimensional SE function,
\begin{equation}\label{eq:TimeSE}
T_i(t,t') = \phi^2\exp\left(\frac{-|t - t'|^2}{2\tau^2}\right).
\end{equation}
Note that $T_i$ includes the parameter $\phi$, which serves to scale
the covariance function $C_{u_i}$. The second option is the Wendland
class of covariance functions \citep{Wendland2005}. Wendland
covariance functions have compact support and hence their
corresponding covariance matrices are sparse. In our analysis, we
exploit this sparsity with the CHOLMOD software package
\citep{Chen2008}. Wendland functions are positive definite in
$\mathbb{R}^d$, and they can be differentiated $k$ times. The form of
the covariance function depends on the choice of $d$ and $k$. We use
$d=1$ since we are describing the temporal covariance of $u_i$. We use
$k=2$, giving samples of $u_i$ continuous velocities and
accelerations. The corresponding Wendland covariance function is
\begin{equation}\label{eq:Wendland} 
T_i(t,t') = \phi^2\left(1 - \frac{|t - t'|}{\tau}\right)^5_+ 
            \left(\frac{8|t - t'|^2}{\tau^2} + \frac{5|t - t'|}{\tau} + 1\right), 
\end{equation} 
where $(t)_+ = \mathrm{min}(0,t)$.
The third option for $T_i$ is the covariance function for integrated
Brownian motion (IBM). IBM is a Gaussian process with zero mean and
its covariance function can be found by integrating the covariance
function for Brownian motion,
\begin{align}\label{eq:IBM}
T_i(t,t') &= \int_0^t \int_0^{t'} \phi^2 \min(s,s') \,ds'\,ds \\
          &= \frac{\phi^2}{2}\min(t,t')^2 \left(\max(t,t') - \frac{1}{3}\min(t,t')\right), \ \ \ t,t' \geq 0.
\end{align}
IBM has been used in the context of Kalman filtering as a
non-parametric model for the time dependence of geophysical signals
\citep[e.g.,][]{Segall1997, McGuire2003, Ohtani2010, Hines2016a}.
Similar to Brownian motion, IBM has a reference time, $t=0$, at which
the process begins. For some geophysical signals, it is appropriate to
have this reference time. For example, if we are trying to identify
postseismic deformation then $t$ should be zero at the time of the
earthquake.  However, if we are interesting in detecting transient
events, where there is no known start time, then IBM may not be an
appropriate prior. Instead, one should prefer an isotropic choice for
$T_i$, meaning that it is invariant to time shifts and can be written
in terms of $|t - t'|$. In the following analysis, we make the choice
that $t$ is zero on the first epoch of $\data_i^*$. Using an earlier
reference time does not change the results discussed in this section.

% Choosing the parameters

% SHOW REML EQUATION IN HERE

We next use the REML method to determine appropriate parameters for
$X_i$ and each of the three choices for $T_i$. we divide the dataset
into seven subsets which are four months long and each centered on the
time of a SSE. The times of the seven SSEs are determined with tremor
records from \cite{Wech2010}. We find the optimal parameters for $T_i$
and $X_i$ for each subset of data. We choose to make each data subsets
four months long because it is long enough to encompass a SSE in the
Pacific Northwest, while it is short enough to still be
computationally tractable. However, four months is too short to
resolve the sinusoids in $\data_i$, and they are omitted from
$\data_i$ in this REML analysis for Pacific Northwest SSEs.

% Results for our parameters

The estimated hyperparameters for $u$ are summarized in Table 1. Based
on the interquartile ranges, the estimated hyperparameters for the SE
and Wendland covariance functions do not vary significantly between
SSEs. This suggests that the median of estimated hyperparameters
should be an appropriate prior model for all Pacific Northwest SSEs.
For the IBM model, there are several anomalously large values for
$\ell$ and $\phi$, which results in large interquartile ranges.
